Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-22 docs/manuscript.tex

\begin{CodeChunk}
\begin{CodeOutput}
* Main project directory
    + FiguresToExtract
        + Paper1_Figure1_trait1.png
        + Paper1_Figure2_trait2.png
        + Paper1_Figure3_trait3.png
        + Paper2_Figure1_trait1.png
        + Paper2_Figure2_trait2.png
        + Paper2_Figure3_trait3.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract
    + 001_Anderson_1935_Fig1.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract
    + caldat
        + 001_Anderson_1935_Fig1
    + 001_Anderson_1935_Fig1.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data
\end{CodeInput}
\begin{CodeOutput}
 filename   group_id         variable       mean      sd  n         r  plot_type
1 001_Anderson_1935_Fig1.png     setosa  Sepal Width (mm) 3.420332 0.4034259 39 0.7524909 scatterplot
2 001_Anderson_1935_Fig1.png     setosa Sepal Length (mm) 4.998267 0.3831294 39 0.7524909 scatterplot
3 001_Anderson_1935_Fig1.png versicolor  Sepal Width (mm) 2.766857 0.3234723 44 0.5157938 scatterplot
4 001_Anderson_1935_Fig1.png versicolor Sepal Length (mm) 5.936769 0.5292535 44 0.5157938 scatterplot
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract
	+ caldat
	  	+ 001_Anderson_1935_Fig1
    + 001_Anderson_1935_Fig1.png
    + 002_Doe_2013_Fig1.png
    + 002_Doe_2013_Fig3.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
data
               filename        group_id         variable     mean        sd   n         r    plot_type
1 001_Anderson_1935_Fig1.png     setosa  Sepal Width (mm) 3.420332 0.4034259 39 0.7524909 scatterplot
2 001_Anderson_1935_Fig1.png     setosa Sepal Length (mm) 4.998267 0.3831294 39 0.7524909 scatterplot
3 001_Anderson_1935_Fig1.png versicolor  Sepal Width (mm) 2.766857 0.3234723 44 0.5157938 scatterplot
4 001_Anderson_1935_Fig1.png versicolor Sepal Length (mm) 5.936769 0.5292535 44 0.5157938 scatterplot
5      002_Doe_2013_Fig1.png     setosa      Sepal length 5.000336 0.7828656 50        NA  mean_error
6      002_Doe_2013_Fig1.png versicolor      Sepal length 5.931340 1.0466177 50        NA  mean_error
7      002_Doe_2013_Fig1.png viriginica      Sepal length 6.588705 1.2608173 50        NA  mean_error
8      003_Doe_2013_Fig3.png       <NA>      Sepal length 4.948472 0.3624212 50        NA  histogram
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
List of 3
 $ mean_error :List of 1
  ..$ 002_Doe_2013_Fig1.png:'data.frame': 3 obs. of  5 variables:
  .. ..$ id      : Factor w/ 3 levels "setosa","versicolor",..: 1 2 3
  .. ..$ mean    : num [1:3] 5 5.93 6.59
  .. ..$ error   : num [1:3] 0.111 0.148 0.178
  .. ..$ n       : num [1:3] 50 50 50
  .. ..$ variable: chr [1:3] "Sepal length" "Sepal length" "Sepal length"
 $ hist       :List of 1
  ..$ 003_Doe_2013_Fig3.png:'data.frame': 8 obs. of  3 variables:
  .. ..$ midpoints: num [1:8] 4.3 4.5 4.7 4.9 5.1 ...
  .. ..$ frequency: num [1:8] 4 5 7 12 11 6 2 3
  .. ..$ variable : chr [1:8] "Sepal length" "Sepal length" "Sepal length" "Sepal length" ...
 $ scatterplot:List of 1
  ..$ 001_Anderson_1935_Fig1.png:'data.frame':  83 obs. of  8 variables:
  .. ..$ id        : Factor w/ 2 levels "setosa","versicolor": 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ x         : num [1:83] 2.3 2.9 3 3 3 ...
  .. ..$ y         : num [1:83] 4.5 4.4 4.41 4.3 4.8 ...
  .. ..$ group     : num [1:83] 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ col       : Factor w/ 2 levels "red","green": 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ pch       : num [1:83] 19 19 19 19 19 19 19 19 19 19 ...
  .. ..$ y_variable: chr [1:83] "Sepal length (mm)" "Sepal length (mm)" "Sepal length (mm)" "Sepal length (mm)" ...
  .. ..$ x_variable: chr [1:83] "Sepal width (mm)" "Sepal width (mm)" "Sepal width (mm)" "Sepal width (mm)" ...
\end{CodeOutput}
\end{CodeChunk}

@@

\section{Testing}
%% simulated data
%% x people doing y plots

We evaluated the consistency of digitisation using the \pkg{metaDigitise} by simulating a series of data and using these data in each of the four different plot types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice and figures were modified slightly to give users a sense that they were digitising new data. We randomised the order of these same 14 figures (using the same data), and provdied the entire figure directory to a total of 15 indpendent digitisers. Digitisers variaed in their level of experience from those having done meta-analyses or comparative work to those not having any experience doing so. We got users to digitise all 14 figures and then send the data on these to evaluate interobserver variablity for each figure type and to assess the accuracy and differences in digitising data from the same data across different figure types.

@@

%\pkg{metaDigitise} is available through GitHub. WHile it is incredibly flexible already, our future plans include building in abilities to deal with log-transformed axes, including arguments for calculating standard deviations from 95\% confidence intervals and standard errors using the t-distribution to correct the t-value for small sample sizes (currently assumed t-values are 1.96 as is normal), dealing with asymmetric error bars and the possibility of zooming in plots such that greater accuracy can be achieved when digitising. We hope to also provide options for assessing inter-observer reliability in the future.

@@

\section*{Acknowledgments}
We thanks the I-DEEL group at UNSW for increadibly useful feedback, and a host of colleagues for testing, providing feedback and digitising including: Rose O'Dea, Fonti Kar, Malgorzata Lagisz, Julia Riley, Diego Barneche, Erin Macartney, Ivan Beltran, Gihan Samarasinghe, Dax Kellie. DWAN was supported by an Australian Research Council Discovery Early Career Research Award (DE150101774) and UNSW Vice Chancellors Fellowship and SN an Australian Research Council Future Fellowship (FT130100268).

>>>>

\begin{CodeChunk}
\begin{CodeOutput}
* Main project directory
    + FiguresToExtract/
        + Paper1_Figure1_trait1.png
        + Paper1_Figure2_trait2.png
        + Paper1_Figure3_trait3.png
        + Paper2_Figure1_trait1.png
        + Paper2_Figure2_trait2.png
        + Paper2_Figure3_trait3.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract/
    + 001_Anderson_1935_Fig1.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract/
    + caldat/
        + 001_Anderson_1935_Fig1
    + 001_Anderson_1935_Fig1.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data
\end{CodeInput}
\begin{CodeOutput}
                  filename   group_id         variable  mean  sd  n    r  plot_type
001_Anderson_1935_Fig1.png     setosa  Sepal Width (mm) 3.42 0.40 39 0.75 scatterplot
001_Anderson_1935_Fig1.png     setosa Sepal Length (mm) 5.00 0.38 39 0.75 scatterplot
001_Anderson_1935_Fig1.png versicolor  Sepal Width (mm) 2.77 0.32 44 0.52 scatterplot
001_Anderson_1935_Fig1.png versicolor Sepal Length (mm) 5.94 0.53 44 0.52 scatterplot
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
*FiguresToExtract/
	+ caldat/
	  	+ 001_Anderson_1935_Fig1
    + 001_Anderson_1935_Fig1.png
    + 002_Doe_2013_Fig1.png
    + 002_Doe_2013_Fig3.png
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
data
               filename      group_id         variable  mean  sd   n    r   plot_type
001_Anderson_1935_Fig1.png     setosa  Sepal Width (mm) 3.42 0.40 39 0.75 scatterplot
001_Anderson_1935_Fig1.png     setosa Sepal Length (mm) 5.00 0.38 39 0.75 scatterplot
001_Anderson_1935_Fig1.png versicolor  Sepal Width (mm) 2.77 0.32 44 0.52 scatterplot
001_Anderson_1935_Fig1.png versicolor Sepal Length (mm) 5.94 0.53 44 0.52 scatterplot
     002_Doe_2013_Fig1.png     setosa      Sepal length 5.00 0.78 50   NA  mean_error
     002_Doe_2013_Fig1.png versicolor      Sepal length 5.93 1.05 50   NA  mean_error
     002_Doe_2013_Fig1.png viriginica      Sepal length 6.59 1.26 50   NA  mean_error
     003_Doe_2013_Fig3.png       <NA>      Sepal length 4.95 0.36 50   NA  histogram
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
List of 3
 $ mean_error :List of 1
  ..$ 002_Doe_2013_Fig1.png:'data.frame': 3 obs. of  5 variables:
  .. ..$ id      : Factor w/ 3 levels "setosa","versicolor",..: 1 2 3
  .. ..$ mean    : num [1:3] 5 5.93 6.59
  .. ..$ error   : num [1:3] 0.111 0.148 0.178
  .. ..$ n       : num [1:3] 50 50 50
  .. ..$ variable: chr [1:3] "Sepal length" "Sepal length" "Sepal length"
 $ hist       :List of 1
  ..$ 003_Doe_2013_Fig3.png:'data.frame': 8 obs. of  3 variables:
  .. ..$ midpoints: num [1:8] 4.3 4.5 4.7 4.9 5.1 ...
  .. ..$ frequency: num [1:8] 4 5 7 12 11 6 2 3
  .. ..$ variable : chr [1:8] "Sepal length" "Sepal length" ...
 $ scatterplot:List of 1
  ..$ 001_Anderson_1935_Fig1.png:'data.frame':  83 obs. of  8 variables:
  .. ..$ id        : Factor w/ 2 levels "setosa","versicolor": 1 1 1 1 1 ...
  .. ..$ x         : num [1:83] 2.3 2.9 3 3 3 ...
  .. ..$ y         : num [1:83] 4.5 4.4 4.41 4.3 4.8 ...
  .. ..$ group     : num [1:83] 1 1 1 1 1 1 1 1 1 1 ...
  .. ..$ col       : Factor w/ 2 levels "red","green": 1 1 1 1 1 1 1 1 1 ...
  .. ..$ pch       : num [1:83] 19 19 19 19 19 19 19 19 19 19 ...
  .. ..$ y_variable: chr [1:83] "Sepal length (mm)" "Sepal length (mm)"  ...
  .. ..$ x_variable: chr [1:83] "Sepal width (mm)" "Sepal width (mm)"    ...
\end{CodeOutput}
\end{CodeChunk}

@@

\section{Interobserver Variability and Validation}
%% simulated data
%% x people doing y plots
\subsection{Methods}
We evaluated the consistency of digitisation using \pkg{metaDigitise} and tested how well digitisations estimate the true underlying data. This was accomplished by simulating a dataset with two different groups and two traits, and these data were then used to construct four different plot types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice for a given plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. We randomised the order of these figures (using the same data) and provdied the entire figure directory to a total of 15 indpendent digitisers. The same data was presented twice for each figure type to estimate between- and within- user variability (i.e., the intra-class correlation coefficient) in the data. Digitisers varied in their level of experience from those having done meta-analyses or comparative work to those not having any experience doing so. We got users to digitise all 14 figures and then send the data on these to evaluate interobserver variablity.

We analysed these data using linear mixed effect models with user identify as a random effect. The overall model also included plot type to account for the different means across the variables and plots. We used the mean, standard deviation and correlation coeficients as response variables in three seperate models. If digitisations were consistent across all users then we would predict no significant between user variability in the data. We also qualitatively report how close the average digitised mean, SD and correlation coefficient were to the actual mean values of traits presented in the figures.

\subsection{Results}

@@

@@

\section*{Acknowledgments}
We thank the I-DEEL group at UNSW for increadibly useful feedback, and a host of colleagues for testing, providing feedback and digitising including: Rose O'Dea, Fonti Kar, Malgorzata Lagisz, Julia Riley, Diego Barneche, Erin Macartney, Ivan Beltran, Gihan Samarasinghe, Dax Kellie, Jonathan Noble and Yian Noble. DWAN was supported by an Australian Research Council Discovery Early Career Research Award (DE150101774) and UNSW Vice Chancellors Fellowship and SN an Australian Research Council Future Fellowship (FT130100268).
